---
title: "Give your paper a descriptive title"
author: "Avery Hammond"
abstract: \singlespacing 1-2 sentences max setting up why your paper is important.
  One sentence that describes your research question. 1-2 sentences describing your
  data and methods. 1-2 sentences describing your main results. One sentence stating
  the implications (optional).
thanks: Use this space to acknowledge help you received completed the project. You
  should thank your peer reviewer and your code checker by name and acknowledge comments
  from your classmates in ECO 324 at Smith College Spring 2020, in addition to anyone
  else that provided useful assistance.
header-includes:
- \usepackage{setspace}
- \doublespacing
output:
  pdf_document:
    number_sections: yes
bibliography: template.bib
---

```{r setup, echo = F, message = F}
knitr::opts_chunk$set(results = 'asis', cache = F)
library(tidyverse)
library(summarytools)
library(stargazer)

```


# Introduction

Follow the introduction "formula" we developed in class to write your introduction. A well written introduction is relatively short and accomplishes the following tasks:
 TALK ABOUT NITROUS OXIDES, HEALTH EFFECTS, ETC
- Motivates the importance of the research question
- Clearly states the research question, the methods, and the main findings

# Literature Review

This paper expands on much of the existing literature regarding the impact of public transit expansion on local air quality in urban settings. Many recent studies have found a correlation between subway expansion and air quality improvement (eg. Titos et al., 2015, Bel et al., 2018, Zheng et al., 2019, Li et al., 2019) Many papers have analyzed the impact of public buses in particular, with Titos et al. (2015) finding black carbon and PM10 reductions of 37% and 33% respectively after the implementation of a new public bus system in Granada, Spain. Bel et al. (2018) have also found a 7.17% reduction in the short-term of air pollution concentrations in the areas surrounding Mexico City's new Metrobus line. This paper varies from these works both in the type of public transit and the length of time analyzed. While users of public transit often utilize both buses and rail systems in their commutes, this study focuses specifically on lightrail systems, such as in Houston, which allow for higher occupancy than individual buses. This study further extends the window of analysis to one year before and after the introduction of the METRO Rail to better estimate the medium-term effects of lightrail expansion. 

We base much of our method on the works of Zheng et al. (2019), and Li et al. (2019),  both of which use difference-in-difference models. Zheng et al. (2019) examines the impact of a new subway line in Changsha, China, on individual pollutant concentrations within areas surrounding the new subway stations. Their difference-in-difference model also includes one year before and after the subway expansion, which this study mirrors both to measure the medium-term effects on air pollution in the affected areas and to avoid seasonal variation in air quality that may impact results. Zheng et al. (2019) find an 18.1% greater reduction in carbon monoxide in areas close to the new subway line compared to more remote areas, but find no significant impact on PM_2.5 or O_3. Li et al. (2019) also use a difference-in-difference model to estimate the impact of an increase in subway density on local air quality in Beijing, China. The authors use both the API (Air Pollution Index) and AQI (Air Quality Index) in their analysis and find that a one standard deviation increase in subway density results in a 1.5% decrease of air pollution concentration in the long-term. This paper applies a difference-in-difference model similar to both Li et al. (2019) and Zheng et al. (2019), weighted for distance between air quality monitors and new subway station locations. 

# Data and Methods

```{r loading data}
load("processed-data/analysis_data.RData")
```


## A Theoretical Model of NO_X Concentrations

As evidenced in the literature review, the introduction of public transit, particularly buses and subways, has been found to reduce concentrations of certain air pollutants in major cities across the world. Some studies (eg. Titos et al., 2015, Zheng et al., 2019) attribute the improvement in air quality to a decrease in traffic congestion, as access to public transit may reduce the need for single-occupancy vehicle travel and therefore decrease vehicle emissions. Others (Li et al., 2019) estimate the cost-benefits of subway expansion and the public health impact of air quality improvement. This study focuses on the nitrogen oxides pollutant group, which are products of vehicle emissions and known respiratory irritants. Because exposure to nitrogen oxides can result in long-term health effects, its inclusion in air quality analyses is highly relevant [CITE NIH SOURCE HERE]. Although I do not assume the findings in this study to be applicable to all U.S. cities, these results may provide a general notion of how nitrogen oxides in particular are impacted by lightrail transit expansion in cities similar to Houston.


## Data

The time period analyzed in this study spans from 2003 to 2004 in Houston, Texas metropolitan area. Time-series data covering all hourly weather measurements and pollutant concentrations and cross-sectional data containing the locations of air quality monitors are retrieved from the Texas Commission on Environmental Quality (Texas Commission on Environmental Quality, 2003). This data was used to create the dummy variable "after," which is equal to 1 for all hourly nitrogen\_oxides and weather measurements after the introduction of the METRO Rail system on January 1, 2004. 

This study also uses cross-sectional data on the locations of Houston Metro stations to analyze the Metro's impact on air quality in these areas, retrieved from the City of Houston GIS database (City of Houston, 2019). I use the coordinates of the new METRO Rail stations and the site data of air quality monitors to compute the distances between each METRO station and the city's air quality monitors. From this, the "exposed" dummy variable was created for air quality monitors within 10 kilometers of a new METRO station. 

\begin{table}
\centering
\caption{Variable definitions}
\label{variableDefinitions}
\begin{tabular}{ll}
\hline
\hline
Variable name      & Description                                                                \\    
\hline
exposed            & Air quality monitors located within 3 kilometers of a new subway station   \\
after              & Dates following the introduction of the METRO Rail on 01/01/2004           \\
ozone              & Concentration of ozone in parts per billion                                \\
nitrogen$_oxides   & Concentration of nitrogen oxides in parts per billion                      \\       
TMP1               & Temperature measured in degrees celcius                                    \\
WDR1               & Wind direction measured in degrees compass                                 \\
WSR1               & Wind speed measured in meters per second                                   \\
\hline
\end{tabular}
\end{table}
 
Following the guidelines in the assignment prompt, describe the key features of your data. This section should include a table of summary statistics for all numerical variables in your data and a discussion of important patterns or caveats.

```{r creating nox summary table}

nox_summary_data <- nox_regression_data %>%
   select(nitrogen_oxides, exposed, after, WSR1, TMP1, WDR1)

stargazer(nox_summary_data, type = "latex",
          label = "SummaryStats",
          title = "Summary Statistics",
          header = F,
          digits = 2)

```

Table \ref{SummaryStats} displays the summary statistics for the explanatory and independent variables. The range in hourly concentrations of nitrogen\_oxides is especially notable, spanning from 0 to 809 parts per billion. Although this maximum value is likely an outlier when compared to the 75th percentile and the mean, such concentrations are highly toxic to humans even for a short period. For the most part, however, the concentrations of nitrogen\_oxides appear to be well within the EPA's 1-hour Primary National Ambient Air Quality Standards of 100 parts per billion. Since "after" and "exposed" are both dummy variables, their summary statistics are not particularly relevant. 

## Model Specification

$nitrous\_oxides_{it} = \alpha_{0} + \beta_{0} + \beta_{1}after + \beta_{2}exposed + \beta_{3}after*exposed + \beta_{4}hour + \beta_{5}TMP1_{it} +  \beta_{6}WSR1_{it} + \beta_{7}WDR1_{it} + \epsilon_{it}$

where $i$ indexes air quality monitors and $t$ indexes hour and date. The hour variable is included as a fixed effect to control for time of day, as pollutant concentrations are expected to vary throughout a 24-hour period. TMP1, WSR1, and WDR1 are included as fixed effects because temperature, wind speed, and wind direction are known to have significant impacts on air quality and pollution levels. These variables also help control for weather variation throughout the year that may cause exogenous fluctuations in pollutant concentrations. The coefficient of after*exposed will be most relevant in determining the effect of lightrail system expansion on nitrous\_oxides concentrations, as it will apply only to measurements taken by air quality monitors within a 10 kilometer radius of a new METRO station during the station's first year of operation. Its coefficient will therefore inform the degree to which the concentrations of nitrous\_oxides changed over this time period as a result of exposure to the new METRO Rail system.


# Results & Discussion

```{r running nox regression}

nox_regression_results <- lm(nitrogen_oxides ~ after + exposed + exposed*after + as.factor(hour) + TMP1 + WSR1 + WDR1, 
                         data = nox_regression_data)

stargazer(nox_regression_results, type = "latex",
          report = 'vctp',
          intercept.bottom = F,
          header = F,
          digits = 5)
```

Present your results here and discuss them. Make sure to consider the magnitude of your effects. This requires you to understand your data. Consider using the standard deviation of your variables to give readers a sense of the size of the effects.

Make sure you return to your research question and discuss what your results tell us about the question you set out to answer.

*Note: if you are using RMarkdown to write your paper, I encourage you to conduct your analysis directly within this file.*

# Conclusion

This brief section will summarize the main conlcusios of the paper, offer concluding comments and may include some discussion of caveats, possible next steps for someone interested in your research *topic*, and any policy or other implications you wish to discuss.

# References
