---
title: "The Impact of Public Transit Introduction on Individual Air Pollutants in Houston, Texas"
author: "Avery Hammond"
abstract: \singlespacing 1-2 sentences max setting up why your paper is important.
  One sentence that describes your research question. 1-2 sentences describing your
  data and methods. 1-2 sentences describing your main results. One sentence stating
  the implications (optional).
thanks: Special thanks to Professor Susan Sayre for advising and coding assistance, Shengqi Zhong for peer review, and comments from Abby Hanna, Kate Ginder, Katya Garcia-Israel, and Victoria Tse in ECO 324 at Smith College Spring 2020.
header-includes:
   - \usepackage{booktabs}
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage[labelfont=bf]{caption}
output:
  pdf_document:
    number_sections: yes
---

```{r setup, echo = F, message = F}
knitr::opts_chunk$set(results = 'asis', cache = F, echo = F)
library(tidyverse)
library(summarytools)
library(stargazer)

```


# Introduction

Follow the introduction "formula" we developed in class to write your introduction. A well written introduction is relatively short and accomplishes the following tasks:
- Motivates the importance of the research question
- Clearly states the research question, the methods, and the main findings

Activity in the transportation sector is regarded around the world as one of the major causes of air pollution, particularly in urban areas. High volumes of vehicles using limited road space causes severe traffic congestion in cities across the globe, which further adds to air pollution in these areas. Drivers in Houston, ranked the 11th most congested city in the U.S. by one analysis, spent 81 hours in congestion in 2019 alone (INRIX, 2020)

As many cities around the world continue to expand, high volumes of vehicles using limited road space results in severe traffic congestion, further adding to

In 2018, transportation comprised 28% of all U.S. greenhouse gas emissions, more than any other sector (US EPA, 2017). 



# Literature Review

This paper expands on much of the existing literature regarding the impact of public transit expansion on local air quality in urban settings. Many recent studies have found a correlation between subway expansion and air quality improvement (eg. Titos et al., 2015, Bel et al., 2018, Zheng et al., 2019, Li et al., 2019) Many papers have analyzed the impact of public buses in particular, with Titos et al. (2015) finding black carbon and PM10 reductions of 37% and 33% respectively after the implementation of a new public bus system in Granada, Spain. Bel et al. (2018) have also found a 7.17% reduction in the short-term of air pollution concentrations in the areas surrounding Mexico City's new Metrobus line. This paper varies from these works both in the type of public transit and the length of time analyzed. While users of public transit often utilize both buses and rail systems in their commutes, this study focuses specifically on light-rail systems, which allow for higher occupancy than individual buses. This study further extends the window of analysis to one year before and after the introduction of the METRO Rail to better estimate the medium-term effects of light-rail expansion in Houston. Many papers (Zheng et al, 2019, etc.) theorize that the improvement in air quality results from the transition from private to public transit, and one recent study has found that subway proximity in Beijing reduces a household's probability of owning a vehicle and consuming fuel (Zhang et al., 2017). This paper attempts to determine the presence of that transition from private to public transport by examining how specific, individual vehicle pollutants are impacted by public transit introduction in order to better estimate the cause of the change.

Much of the method used in this study is based on the works of Zheng et al. (2019), and Li et al. (2019), both of whom use difference-in-difference models. Zheng et al. (2019) examines the impact of a new subway line in Changsha, China, on individual pollutant concentrations within areas surrounding the new subway stations. Their difference-in-difference model also includes one year before and after the subway expansion, which this study mirrors both to measure the medium-term effects on air pollution in the affected areas and to avoid seasonal variation in air quality that may impact results. Zheng et al. (2019) find an 18.1% greater reduction in carbon monoxide in areas close to the new subway line compared to more remote areas, but find no significant impact on $PM_{2.5}$ or $O_3$. Li et al. (2019) also use a difference-in-difference model to estimate the impact of an increase in subway density on local air quality in Beijing, China. The authors use both the API (Air Pollution Index) and AQI (Air Quality Index) in their analysis and find that a one standard deviation increase in subway density results in a 1.5% decrease of air pollution concentration in the long-term. The authors do not include individual pollutants in their regression, but rather overall air quality indicators. This paper applies a difference-in-difference model similar to both Li et al. (2019) and Zheng et al. (2019), weighted for distance between air quality monitors and new subway station locations. 

# Data and Methods

```{r loading data}
load("processed-data/analysis_data.RData")
```


## A Conceptual Model of Vehicle Pollutant Concentrations

As evidenced in the literature review, the introduction of public transit, particularly buses and subways, has been found to reduce concentrations of certain air pollutants in major cities across the world. Very few of these studies, however, have been conducted within the U.S. Some studies (eg. Titos et al., 2015, Zheng et al., 2019) attribute the improvement in air quality to decreased traffic congestion as individuals transition from private to public transport, which may reduce single-occupancy-vehicle emissions. This study looks specifically at four major vehicular pollutants: $NO_{x}$, $CO$, $SO_{2}$, and $PM_{2.5}$. The analyses of these distinct pollutants are necessary in order to examine the relationship between vehicle-induced air pollution and nearby public transport expansion, because overall air quality indicators may be influenced by non-vehicular emissions. By analyzing pollutants individually, I attempt to examine precisely how air quality changes with local introduction of public transit in Houston, as some pollutants may change significantly and others not at all. Since each of these pollutants pose different threats to public health and the environment, the distinction between them is highly relevant. I aim to determine overall whether the introduction of the Houston METRO Rail resulted in nearby vehicle pollutant concentration reductions in the first year of the light-rail system's operation. Although I do not assume the findings in this study to be applicable to all U.S. cities, these results may provide a general notion of how vehicle pollutants can be impacted by light-rail transit expansion in cities similar to Houston.


## Data

The time period analyzed in this study spans from 2003 to 2004 in Houston, Texas metropolitan area. Time-series data covering all hourly weather measurements and pollutant concentrations and cross-sectional data containing the locations of air quality monitors are both retrieved from the Texas Commission on Environmental Quality (Texas Commission on Environmental Quality, 2003). This data was used to create the dummy variable "after," which is equal to 1 for all hourly pollutant concentrations and weather measurements after the introduction of the METRO Rail system on January 1, 2004. I also altered the original wind direction variable to correspond to compass directions instead of compass degrees to account for the circular nature of directional measurement (e.g. 1° and 359° are now both categorized as North).

This study also uses cross-sectional data on the locations of Houston Metro stations to analyze the Metro's impact on air quality in these areas, retrieved from the City of Houston GIS database (City of Houston, 2019). I use the coordinates of the new METRO Rail stations and the site data of air quality monitors to compute the minimum distances between each METRO station and the city's air quality monitors. From this, the "exposed" dummy variable was created for air quality monitors within 10.5 kilometers of a new METRO station. The distance of 10.5 kilometers was chosen in order to include air quality monitors that measured $PM_{2.5}$ and $CO$, none of which were located closer than 10.4 kilometers away from a new METRO station. It is also noteworthy that not all air quality monitors measure every pollutant used in this study, which accounts for the variation in N-values in the summary tables. Table \ref{variabledefinitions} provides definitions for the variables used in this analysis.

\begin{table}
\centering
\caption{Variable Definitions}
\label{variabledefinitions}
\begin{tabular}{ll}
\hline
\hline
Variable name      & Description                                                                          \\    
\hline      
exposed            & Air quality monitors located within 10.5 kilometers of a new subway station          \\
after              & Dates following the introduction of the METRO Rail on 01/01/2004 until 12/31/2004    \\
carbon\_monoxide   & Hourly concentration of carbon monoxide in parts per billion                         \\
log\_co            & Log of carbon\_monoxide                                                              \\
nitrogen\_oxides   & Hourly concentration of nitrogen oxides in parts per billion                         \\ 
log\_nox           & Log of nitrogen\_oxides                                                              \\
sulfur\_dioxide    & Hourly concentration of sulfur dioxide in parts per billion                          \\
log\_so2           & Log of sulfur\_dioxide                                                               \\
PM\_25             & Hourly concentration of $PM_{2.5}$ in micrograms per cubic meter                     \\ 
log\_pm25          & Log of PM\_25                                                                        \\
TMP1               & Temperature measured in degrees celsius                                              \\
WDR                & Wind direction measured in compass directions                                        \\
WSR1               & Wind speed measured in meters per second                                             \\
\hline
\end{tabular}
\end{table}
 

```{r creating nox summary table}

nox_summary_data <- nox_regression_data %>%
   select(nitrogen_oxides, exposed, after, WSR1, TMP1, WDR)

stargazer(nox_summary_data, type = "latex",
          label = "NOXSummaryStats",
          title = "$NO_{x}$ Summary Statistics",
          header = F,
          digits = 2)

```

```{r creating co summary table}

co_summary_data <- co_regression_data %>%
   select(carbon_monoxide, exposed, after, WSR1, TMP1, WDR)

stargazer(co_summary_data, type = "latex",
          label = "COSummaryStats",
          title = "$CO$ Summary Statistics",
          header = F,
          digits = 2)

```

```{r creating so2 summary table}

so2_summary_data <- so2_regression_data %>%
   select(sulfur_dioxide, exposed, after, WSR1, TMP1, WDR)

stargazer(so2_summary_data, type = "latex",
          label = "SO2SummaryStats",
          title = "$SO_{2}$ Summary Statistics",
          header = F,
          digits = 2)

```

```{r creating PM2.5 summary table}

pm25_summary_data <- pm25_regression_data %>%
   select(PM_25, exposed, after, WSR1, TMP1, WDR)

stargazer(pm25_summary_data, type = "latex",
          label = "PM25SummaryStats",
          title = "$PM_{2.5}$ Summary Statistics",
          header = F,
          digits = 2)

```

The tables above display the summary statistics for the explanatory and each independent pollutant variable. It is pertinent to reiterate that the units of nitrogen oxides, sulfur dioxide, and carbon monoxide in this study are all measured in parts per billion (ppb). 1-hour National Ambient Air Quality Standards (NAAQS) for carbon monoxide are set in parts per million (ppm), and in ppb for both nitrogen dioxide and sulfur dioxide (US EPA, 2014). $PM_{2.5}$ does not have a 1-hour exposure standard set by the EPA and is measured in micrograms per cubic meter both in this study and in regulatory standards. The average 1-hour concentration of $carbon\_monoxide$ as seen on this table is equivalent to 0.47399 ppm, which is well within the National Ambient Air Quality Standards of 9ppm (US EPA, 2014). The hourly means of all pollutants appear to stay within a relatively low range from 2003 to 2004 and all are within National Ambient Air Quality Standards set by the EPA (US EPA, 2014). 

The range in hourly concentrations of $nitrogen\_oxides$ is especially notable, spanning from 0 to 809 ppb. Since nitrogen oxides are a group of gases, it is difficult to tell which specific gas contributed to this maximum. It is notable however that the EPA has set the 1-hour National Ambient Air Quality Standard of nitrogen dioxide, the most common nitrogen oxide, to 100 ppb (US EPA, 2016). Although this maximum value is likely an outlier when compared to the 75th percentile and the mean, such concentrations of nitrogen dioxide would be toxic to humans even for a short period (US EPA, 2016). The maximum value of hourly sulfur dioxide concentrations also exceed the EPA's air quality standards of 75ppb, but the mean values and percentiles would indicate that this measurement is also an outlier. From these maximums, however, it is evident that the data is skewed towards higher concentration values. 

The mean values of $exposed$ show that 16% of air quality stations that measured $NO_x$ were within 10.5 km, as were 33% of monitors that measured $CO$, 60% of monitors that measured $SO_2$, and 12% of monitors that measured $PM_{2.5}$.

## Model Specification

The dependent variable in this regression is the log of the pollutant ($NO_x$, $CO$, $SO_2$, or $PM_{2.5}$). This study uses a linear difference-in-difference model, specified as:

$log\_pollutant_{itp} = \beta_{i} + \beta_{1}after + \beta_{2}exposed + \beta_{3}after*exposed + \beta_{4}hour + \beta_{5}TMP1_{it} +  \beta_{6}WSR1_{it} + \beta_{7}WDR_{it} + \epsilon_{it}$

where $p$ indexes the individual pollutant, $i$ indexes air quality monitors, and $t$ indexes hour and date. $\beta_{i}$ represents the air quality monitor fixed effect. The log of the pollutants are used to help account for skewed data, the presence of which is evident from the maximums in the summary statistics tables. $hour$ is included as a fixed effect to control for time of day, as pollutant concentrations are expected to vary throughout a 24-hour period. $TMP1$, $WSR1$, and $WDR$ are included as fixed effects because temperature, wind speed, and wind direction are known to have significant impacts on air quality and pollution levels. These variables also help control for weather variation throughout the year that may cause exogenous fluctuations in pollutant concentrations. The dummy variable $exposed$ is equal to one if the air quality monitor $i$ is within 10.5km of a new METRO station; $after$ is equal to one if the date is on or after the introduction of the METRO Rail on 01/01/2004. The coefficient of $after*exposed$ will be most relevant in determining the effect of light-rail system expansion on pollutant concentrations, as it will apply only to measurements taken by air quality monitors within a 10 kilometer radius of a new METRO station during the station's first year of operation. Its coefficient will therefore inform the magnitude to which the concentrations of each pollutant changed over this time period as a result of exposure to the new METRO Rail system.


# Results & Discussion


```{r running nox regression}

nox_regression_results <- lm(log_nox ~ after + exposed + exposed*after + as.factor(hour) + TMP1 + WSR1 + WDR, 
                         data = nox_regression_data)

```

```{r running co regression}

co_regression_results <- lm(log_co ~ after + exposed + exposed*after + as.factor(hour) + TMP1 + WSR1 + WDR, 
                         data = co_regression_data)

```

```{r running PM2.5 regression}

pm25_regression_results <- lm(log_pm25 ~ after + exposed + exposed*after + as.factor(hour) + TMP1 + WSR1 + WDR, 
                         data = pm25_regression_data)

```

```{r running so2 regression}

so2_regression_results <- lm(log_so2 ~ after + exposed + exposed*after + as.factor(hour) + TMP1 + WSR1 + WDR, 
                         data = so2_regression_data) 

```

```{r creating a combined table of results}
stargazer(nox_regression_results, co_regression_results, pm25_regression_results, so2_regression_results,
          type = "latex",
          title = "Regression Results",
          label = "regressionresults",
          header = F,
          digits = 2,
          omit = c("hour", "WDR"),
          column.sep.width = "0pt",
          font.size = "scriptsize",
          single.row = FALSE,
          notes = "Hour fixed effects and wind direction fixed effects also included but coefficients are omitted here.",
          notes.align = "l")

```

## Results

The results of all pollutant regressions can be found on Table \ref{regressionresults}. With the exception of $PM_{2.5}$, all coefficients of $after*exposed$ are statistically significant. These coefficients suggest that in the year following Houston METRO Rail expansion, air quality monitors inside a 10.5 kilometer radius of new METRO stations experienced subsequent concentration reductions in nitrogen oxides and carbon monoxide by 8.61% and 36.14% respectively, while sulfur dioxide concentrations experienced a 7.22% increase. As I hypothesized in the Data and Methods section, individual pollutants experienced varying levels of reduction in the year following the introduction of the METRO Rail, with the difference in $PM_{2.5}$ being insignificant entirely. 

## Discussion

The pollutant that experienced by far the largest reduction as a result of subway expansion is carbon monoxide, the greatest source of which is motorized vehicles that burn fossil fuel (US EPA, 2016). These findings resemble those of Zheng et al. (2019) and Bel et al. (2018), both of whom find significant reductions in carbon monoxide in areas located near new or expanded public transit. The greatest source of nitrogen oxides are also motorized vehicles, and both $NO_x$ and $CO$ are primary vehicle pollutants, meaning they are released into the atmosphere directly from the source instead of forming as secondary pollutants through the chemical reactions of other particles (US EPA, 2016). This characteristic, combined with the major cause of both pollutants coming from vehicle emissions, makes nitrogen oxides and carbon monoxide the strongest indicators of a transition from private to public transit in this study. Statistically significant reductions in carbon monoxide and nitrogen oxides further support the findings of Zheng et al. (2019) and many other studies, which theorize that the improvements in air quality as a result of public transport expansion may be caused by the substitution of private vehicles for public, therefore causing fewer emissions. 

As a secondary pollutant, the insignificance of $PM_{2.5}$'s $after*exposed$ coefficient may be due its dependence on sulfur dioxide and nitrogen oxides to form. Because nitrogen oxides decreased and sulfur dioxide increased, there may have been conflicting factors in the formation of $PM_{2.5}$, resulting in an insignificant coefficient. Although sulfur dioxide is a major vehicle pollutant, the primary contributors of atmospheric $SO_2$ are industrial processes, which may explain why sulfur dioxide concentrations in the area increased during the analyzed time period (US EPA, 2016). The low $R^2$ value suggests that the explanatory variables included in the regression only account for about 8% of the variation in sulfur dioxide, which may also indicate that exogenous industrial activity played a role in $SO_2$ concentrations. Low $R^2$ values overall suggest that much of the variation in the pollutant variables is not accounted for by the independent variables used in this regression. Other potentially influential explanatory variables include precipitation and humidity, the data for which was unavailable at the time of this study. The overall results of this regression indicate that the expansion of public transit causes lower nitrogen oxides and carbon monoxide concentrations in at least one U.S. city in the medium-term, and that the findings of similar research in other countries may be applicable to parts of the United States. 


# Conclusion

This study adds to the extensive field of literature that analyzes the relationship between public transit and air pollution to better understand the determinants of air quality in urban settings. The results of this paper indicate that after exposure to local public transport introduction in Houston, concentrations of carbon monoxide decreased by 36.1%, nitrogen oxides decreased by 8.6%, sulfur dioxide increased by 7.2%, and $PM_{2.5}$ experienced no significant change within the first year of the METRO Rail's operation. Because atmospheric sulfur dioxide primarily comes from industrial processes, such a significant increase may caused by industrial activity in the surrounding area rather than as a result of public transit introduction. Additionally, the lack of significance in $PM_{2.5}$ may be attributed to the role of nitrogen oxides and sulfur dioxide in its formation. Because nitrogen oxides and sulfur dioxide experienced changes of similar magnitude in opposite directions, there may have been conflicting factors in the formation of $PM_{2.5}$, resulting in an insignificant result. These conflicting or exogenous factors in sulfur dioxide and $PM_{2.5}$ concentrations, in addition to $NO_x$ and $CO$'s major sources coming from vehicle emissions, makes nitrogen oxides and carbon monoxide the strongest indicators of a transition from private to public transport out of the pollutants in this study (US EPA, 2016). The results of the DID model used in this paper support many other works that find a similar reduction in carbon monoxide concentrations as a result of proximity to new public transport, in addition to providing further evidence on the magnitude and direction of changes in individual pollutant concentrations after exposure to nearby public transit introduction. Potential steps for future research on this topic include the addition of precipitation and humidity variables, the exploration of conflicting causes in different pollutant sources, and the examination the distinction between public transit expansion and public transit introduction.

# References

[Will insert later]
